\usefont{OT1}{bch}{b}{n} \normalfont \normalsize \textsc{University of Queensland} \\ [25pt] \horrule{0.5pt} \\[0.4cm] \huge Epistatic control of gene expression in Humans\\[0.3cm] \huge *** Report V1 *** \\ \horrule{2pt} \\[0.5cm]

### 2013-04-24 docs/manuscript/report_v1.tex
added some text on the simulations / fine mapping. Need to wait until some more analyses finish before I can include full results here

To assess the impact of pair-wise epistasis variants not present on the Illumina 610Q beadchip and better define the association regions, we imputed genotypes in the region 100kb upstream and downstream of the sentinel SNPs using the full reference panel from the 1000 Genomes Project. The quality of imputation was assessed by means of the $R^2$ statistic. Used a highly conservative threshold, we removed any SNPs with $R^2 < 0.9$. Thus, or each epistasis pair two imputed SNP blocks were carried forward for fine mapping analysis. Each pairwise combination of SNPs in blocks 1 and 2 were tested in a ful 8d.f. and reduced (interaction) 4d.f. model. As before any snp pairs not matching the criteria of $r^2$  < 0.1, number of genotype classes = 9, and minimum genotype class size > 5 were removed from the output.

This approach allows us to fine map to the causal epistatic loci. Given the nature epistasis genetic variance, any snp pairs with large estimate of non-additive genetic variance are very unlikely to be rare. Therefore, fine mapped loci provide good functional candidates. However, as fine mapping includes a large number of additional tests increases in the interaction (4d.f.) $p$-values could be due to chance [need a better way of describing this].

>>>>

To assess the impact of pair-wise epistasis variants not present on the Illumina 610Q beadchip and better define the association regions, we imputed genotypes in the region 100kb upstream and downstream of the sentinel SNPs using the full reference panel from the 1000 Genomes Project. The quality of imputation was assessed by means of the $R^2$ statistic. Used a highly conservative threshold, we removed any SNPs with $R^2 < 0.9$. Thus, or each epistasis pair two imputed SNP blocks were carried forward for fine mapping analysis. Each pairwise combination of SNPs in blocks 1 and 2 were tested in a ful 8d.f. and reduced (interaction) 4d.f. model. As before any SNP pairs not matching the criteria of $r^2$  < 0.1, number of genotype classes = 9, and minimum genotype class size > 5 were removed from the output.

This approach allows us to fine map to the causal epistatic loci. Given the nature epistasis genetic variance, any snp pairs with large estimate of non-additive genetic variance are very unlikely to be rare. Therefore, fine mapped loci provide good functional candidates. Hereon, results from the original 2D scan using genotyped SNPs will be refered to as genotype pairs and results from the fine mapping 2D scan using imputed data will be termed imputed pairs. Thus, for each significant pair there will be a results from genotyped and imputed pairs. However, as fine mapping includes a large number of additional tests, any oberved increases in the interaction (4d.f.) $p$-values could be due to chance arising from multiple testing. To address this we ran a series of simulations to determine the empirical change from genotype interaction $p$-value to imputed interaction $p$-value expected in situations where no epistatic exists. For the following three scenarios; $cis-cis$, $cis-trans$ and $trans-trans$ we randomly selected a probe for which no significant epistatic genotype pairs had been identified and randomly selected two SNP blocks within the defined genomic regions. The SNP pair was tested using the 4 and 8d.f. models and then imputed SNPs were extracted from 100kb upstream and downstream regions. Pairwise interactions were tested for the SNPs in the two imputed blocks and the SNP pair with the minimum interaction $p$-value retained. This process was repeated 10,000 times for each of the three scenarios.
